---
title: "Introduction to Quarto"
format: html
bibliography: references.bib
---

Markdown formatting:

Above is called the YAML. This includes data such as title, format, and other data such as date, author, etc.

Here are some examples of things you can do with the formatting...

# This is a large title

## This is a smaller title

####### This is even smaller

*Italic Text* **Bold Text**

```         
Markdown quick reference in the help tab will show what kinds of things you can do and their shortcuts.
```

> Hello Yall Whyyyyy

------------------------------------------------------------------------

```         
It is easier to write citations in the visual tab.
Use the insert tab to onclude a citation. 
It will pop up a seperate tab to search for an article, such as searching for a doi.
Can link to your citation manager if you need it. 

Inserting a table is also easier in the visual tab.
There is a table tab here where you can treat it like a normal text based table. 
Can build then in the source tab, but it's much more complicated to manipulate.
```

Here is an example of a citation. It will provide a link like this and a full citation below.[@oke2020]

Here is an example of a table.

| 1   | 2   | 3   |
|-----|-----|-----|
|     |     |     |
|     |     |     |
|     |     |     |

-----

Working with the BGChem2008 data:

We will be using r functions contained in different R packages for this exercise. 
Later in the course we will learn to write our own R packages.

# 1.Introduction

## a. About the data
  - water sample data from the arctic ocean, downloaded form the arctic data center: [Tweedie ADC data](https://srcticdatacenter.io/catalog/view/doi:10.18739/A25T3FZ8X)

## b.Setup

```{r}
#| warning: false 
library(readr)
library(here)
```
## c. Read in Data
```{r}
#| warning: false
bg_chem <- read_csv(here("data/BGchem2008data.csv"))

```

# 2. Analysis

# a. Calculate summary statistics

```{r}
#| warning: false
#| eval: false
#| echo: false

# FYI the | key is a back slash (above the enter key)
# this code below will list the column names
colnames(bg_chem)

# this will show the data structure
str(bg_chem)

# this will show the first 6 lines of the data frame
head(bg_chem)

# this will show summary statistics of each column
summary(bg_chem)

#this will print unique values in a specific column using the $
unique(bg_chem$Date)

```


# b. Calculate mean Redfield ratio

```{r}
# calculate some means
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)


```


# c. Plot Redfield ratio

```{r}
ratio <- (nitrate + nitrite + amm) / phos
```

Here is the ratio that we calculated: `r round(ratio)`

```
This is a tiny chuck of r we are referring to. Need to put the code in ticks(`)
Anytime you're doing a next step of analysis, it's good to have it in it's own chunk. 
```

Here is a graph of the ratio.

```{r}
plot(bg_chem$P, bg_chem$NO2 + bg_chem$NO3 + bg_chem$NH4)
```




## 3. Conclusion

``` 
It is good to periodically save and render your Quarto, this can help you identify any errors in your code.

Another way to check for errors is to clear your environment completely and run your script from the beginning. This will expose if you have any undefined objects that 
```

# Citations

