---
title: "Tidying and Wrangling"
format: html
---

```{r}
#| message: false

library(dplyr)
library(tidyr)
library(readr)
```

# Data Sources

## Read in our Data

```{r}
#| message: false

catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore

```{r}
# It's a good idea to take a preliminary look at your data first before diving in...

colnames(catch_original)
# column names

head(catch_original)
# provides the first 5 lines of code

summary(catch_original)
# summary statistics of the columns

unique(catch_original$Region)
# looks at the values of specific columns.

```

## The pipe operator %>% (the shortcut is control + shift + m). It will link together different operations. 

# Cleaning and Reshaping

## Using the Select() fucntion

```{r}
# select to keep the columns that you want

catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)

#to drop columns you don't want

catch_data <- catch_original %>% 
  select(-All, -notesRegCode)
head(catch_data)

```

# Check column types
```{r}
# this gives you a look at the data
glimpse(catch_data)

```

## Changing the Column contents with mutate()

```{r}
# change the Chinook column to numeric from characters

catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))

#the error appears because there are some non-numeric values. In our case, a 1 has been replaced with an I. We can fix this.
```

```{r}
is.na(catch_clean$Chinook)

# get the location of the NA value
i <- which(is.na(catch_clean$Chinook))

# look at the row in the data 
catch_data[i,]

# use the mutate function to clean this up
# you can have multiple things you mutate at once in the mutate function.
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(condition = Chinook == "I",
                           true = "1",
                           false = Chinook),
         Chinook = as.numeric(Chinook))

# double check that the error has be resolved 
head(catch_clean)

# you can double check here too
catch_clean[i,]

```

## Changing file shape using pivot_longer() and pivot_wider()

```{r}
# let's put all the species in one column
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), 
               names_to = "species", 
               values_to = "catch")
head(catch_long)

# now lets manipulate it back to what we had before
catch_wide <- catch_long %>% 
  pivot_wider(names_from = species,
              values_from = catch)
head(catch_wide)

#If you 

```

## Rename the Column with the Rename function

```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = catch)


# If you look at the metadata, the catches are actually in repsesenting 1000 catches. Let's fix this.

catch_long <- catch_long %>% 
  mutate(catch = catch_thousands * 1000) %>% 
  select(-catch_thousands)

head(catch_long)

# Now we can work with the data
```

## Summary statistics using group_by() and summarize()

```{r}
#split, combine, apply framework.
# get the average catch by region

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch))
head(mean_region)

# number of rows per each group

n_region <- catch_clean %>% 
  group_by(Region) %>% 
  summarize(n = n())
head(n_region)


```

# Filtering rown using filters()
```{r}
sse_catch <- catch_long %>% 
  filter(Region == "SSE")
head(sse_catch)

#try filtering where catch column is over 1 million fish
#winning, I did it on my own. lol

million_catch <- catch_long %>% 
  filter(catch > 1000000)
head(million_catch)

# you can filter by more than one filters at once

chinook_SSE <- catch_long %>% 
  filter(Region == "SSE" & species == "Chinook")
head(chinook_SSE)

```

## Sorting data with arrange()

```{r}

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(mean_catch)

head(mean_region)

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))
head(mean_region)

```

## Combining datasets together with the join function

```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)

head(region_defs)

catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))
  head(catch_joined)

```

```{r}
region_defs <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea) %>% 
  rename(Region = code, Region_Name = mgmtArea)


catch_joined <- left_join(catch_long, region_defs, by = c("Region"))
head(catch_joined)
  
```

## Splitting and combinin values in columns

```{r}
sites_df <- data.frame(site =c("Haw-101",
                               "Haw-103",
                               "OAH-320"))
head(sites_df)

split_df <- sites_df %>% 
  separate(site, c("island", "site_number"), sep = "-")
head(split_df)

```

```{r}

dates_df <- data.frame(
  year = c("1930", "1930", "1932"),
  month = c("12", "12", "12"),
  day = c("14", "15", "16")
)
head(dates_df)

dates_unite <- dates_df %>% 
  unite(col = date, year, day, sep = "-")
head(dates_unite)


```

```{r}

mean_region <- catch_original %>% 
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I",
                           "1",
                           Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year),
               names_to = "species",
               values_to = "catch") %>% 
  mutate(catch = catch *1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)

```

## save our results with write_csv()

```{r}
library(here)
write_csv(mean_region, here("data/mean_catch_by_region.csv"))


```


